Typed Clojure

### 2014-08-28 schemeworkshop13.tex
add more type rules

\title{Typed Clojure}
\subtitle{A Practical Type System for Clojure, based on Typed Racket}

\authorinfo{Ambrose Bonnaire-Sergeant}
           {University of Western Australia}
           {abonnairesergeant@gmail.com}
\authorinfo{Rowan Davies}
           {University of Western Australia}
           {rowan@csse.uwa.edu.au}

@@

\begin{abstract}
Typed Racket is a valuable starting point for
a gradual type system that targets Clojure.
Building a similar type system for a new language gives the
designer some flexibility to repurpose and extend features.
This paper gives an overview of Typed Clojure, concentrating
on the extensions and differences from Typed Racket. We also
show where Typed Racket's features were particularly useful
for type checking non-trivial Clojure idioms.
\end{abstract}

@@

\section{Typed Clojure}

@@

@@

Typed Clojure is a gradual type system for Clojure, designed to

@@

\begin{lstlisting}[label=lst:ordered]
(Fn [Integer -> Integer]
    [Number -> Number])
\end{lstlisting}

@@

@@

Typed Clojure supports f-bounded polymorphism and higher-rank
types.

@@

\begin{figure*}
$$
\begin{altgrammar}
  \expd{}, \e{} &::=& \x\ \alt {\comb {\e{}} {\e{}}} %\alt {\abs {\x a} {\e{}}}
                      \alt \true{} \alt \false{} \alt \nil{}
                      \alt \class{}
                      \alt {\tryexp {\e{}} {\class} {\x} {\e{}} {\e{}}}
                      \alt {\ifexp {\e{}} {\e{}} {\e{}}}
                      \alt
                      {\doexp {\e{}} {\e{}}}
                      \\ &\alt& 
                      {\fnexp {\overrightarrow{\hastype {\x{}} {\t{}}}} {\e{}}}
                      \alt {\letexp {\x{}} {\e{}} {\e{}}}
                &\mbox{Expressions} \\ \\
  \s{}, \t{}    &::=& \Top \alt (\class{}\ \overrightarrow{\t{}})
                      \alt {\Value \v{}}
                      \alt {\Unionsplice {\overrightarrow{\t{}}}}
                      \alt (\cap\ \overrightarrow{\t{}})
                      \alt (\Not\ \t{})
                      \alt {\All {\overrightarrow{\x{}}} {\t{}}}
                      \alt {\TFn {\overrightarrow{\x{}}} \t{}}
                      \alt {\Arrow {\overrightarrow{ \hastype {\x{}} {\t{}}}}
                                   {\t{}}
                                   {\filterset {\prop{}} {\prop{}}}
                                   {\object{}}
                                   {\prop{}}}
                      \\ &\alt&
                      {\ArrayTwo {\t{}} {\t{}}}
                &\mbox{Types} \\ \\
  \prop{}       &::=& {\isprop {\t{}} {\path {\pathelem{}} {\x{}}}}
                      \alt {\notprop {\t{}} {\path {\pathelem{}} {\x{}}}}
                      \alt {\impprop {\prop{}} {\prop{}}}
                      \alt {\andprop {\prop{}} {\prop{}}}
                      \alt {\orprop {\prop{}} {\prop{}}}
                      \alt \topprop{}
                      \alt \botprop{}
                &\mbox{Propositions} \\
  \object{}     &::=& {\path {\pathelem{}} {\x{}}}
                      \alt \emptyobject{}
                &\mbox{Objects} \\
  \pathelem{}   &::=& \overrightarrow{\pesyntax{}}
                &\mbox{Paths} \\
  \pesyntax{}   &::=& \classpe{}
                &\mbox{Path Elements} \\ \\
                % TODO objects with path
%  \c{} &::=& \assoc{} \alt \dissoc{} \alt \get{}
%              &\mbox{Constants}
  \Gamma       &::=& \overrightarrow{\prop{}}
               &\mbox{Environments}
\end{altgrammar}
$$
\caption{Syntax of Terms, Types, Propositions, and Objects}
\end{figure*}

\begin{figure*}
\begin{mathpar}
\infer [T-Var]
{ \inpropenv {\propenv{}} {\isprop {\x{}} {\t{}}}}
{ \judgement {\propenv{}} {\hastype {\x{}} {\t{}}} {\filterset {{\notprop {\falsy} {\x{}}}} {{\isprop {\falsy} {\x{}}}}}
             {\x{}} {\topprop{}}}

\infer [T-True]
{}
{ \judgement {\propenv{}} {{\hastype {\true{}} {\True{}}}} {{\filterset {\topprop{}} {\botprop{}}}} {\emptyobject{}} {\topprop{}}}

\infer [T-False]
{}
{ \judgement {\propenv{}} {{\hastype {\false{}} {\False{}}}} {\filterset {\botprop{}} {\topprop{}}} {\emptyobject{}} {\topprop{}}}

\infer [T-Nil]
{}
{ \judgement {\propenv{}} {{\hastype {\nil{}} {\Nil{}}}} {\filterset {\botprop{}} {\topprop{}}} {\emptyobject{}} {\topprop{}}}

\infer [T-Do]
{ \judgement {\propenv{}} {\hastype {\e1} {\t1}} {\filterset {\thenprop {\prop{1}}} {\elseprop {\prop1}}} {\object{1}} 
             {\flowprop {\prop{1}}}
\\\\
  \judgement {\propenv{}, {\flowprop {\prop{1}}}} {\hastype {\e2} {\t2}} {\filterset {\thenprop {\prop2}} {\elseprop {\prop2}}} 
             {\object{}} {\flowprop {\prop{2}}}
  }
{ \judgement {\propenv{}} {\hastype {\doexp {\e1} {\e2}} {\t2}} {\filterset {\thenprop {\prop2}} {\elseprop {\prop2}}} {\object} 
             {\andprop {\flowprop {\prop{1}}} {\flowprop {\prop{2}}}}}

\infer [T-If]
{ \judgement {\propenv{}} {\hastype {\e1} {\t{1}}} {\filterset {\thenprop {\prop{1}}} {\elseprop {\prop{1}}}}
             {\object{1}} {\flowprop {\prop{1}}} 
             \\\\
  \judgement {\propenv{}} {\hastype {\e2} {\t{}}} {\filterset {\thenprop {\prop{2}}} {\elseprop {\prop{2}}}}
             {\object{}} {\flowprop {\prop{2}}}
  \\\\
  \judgement {\propenv{}} {\hastype {\e3} {\t{}}} {\filterset {\thenprop {\prop{3}}} {\elseprop {\prop{3}}}}
             {\object{}} {\flowprop {\prop{3}}}
             }
{ \judgement {\propenv{}} {\hastype {\ifexp {\e1} {\e2} {\e3}} {\t{}}} 
            % do we need to include prop1+/- here?
             {\filterset {\orprop {\thenprop {\prop{2}}} {\thenprop {\prop{3}}}} 
                         {\orprop {\elseprop {\prop{2}}} {\elseprop {\prop{3}}}}}
             {\object{}} {\orprop {\flowprop {\prop{2}}}
                                  {\flowprop {\prop{3}}}} }

\infer [T-Let]
{ \judgement {\propenv{}} {\hastype {\e{1}} {\s{}}} {\filterset {\thenprop {\prop{1}}} {\elseprop {\prop{1}}}}
             {\object{1}} {\flowprop {\prop{1}}}
  \\\\
  \judgement {\propenv{}, {\isprop {\s{}} {\x{}}}, 
              {\impprop {\notprop {\falsy} {\x{}}} {\thenprop {\prop{1}}}}, 
              {\impprop {\isprop {\falsy} {\x{}}} {\elseprop {\prop{1}}}}}
             {\hastype {\e{}} {\t{}}} {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
             {\object{}} {\flowprop {\prop{2}}}
             }
{ \judgement {\propenv{}} {\hastype {\letexp {\x{}} {\e{1}} {\e{}}} {\t{}}}
             {\replacefor {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
                          {\object{1}}
                          {\x{}}}
             {\replacefor {\object{}} 
                          {\object{1}}
                          {\x{}}}
             {\replacefor {\andprop {\flowprop {\prop{1}}} {\flowprop {\prop{2}}}}
                          {\object{1}}
                          {\x{}}}}

\infer [T-Throw]
{ \judgement {\propenv{}} {\hastype {\e{}} {\Throwable}} {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
             {\object{}} {\flowprop {\prop{}}}}
{ \judgement {\propenv{}} {\hastype {\throwexp {\e{}}} {\Bottom}} {\filterset {\botprop} {\botprop}}
             {\emptyobject{}} {\botprop} }

\infer [T-Fn]
{}
{ TODO }

\infer [T-New]
{ }
{ \judgement {\propenv{}} {\hastype {\newexp {\class{}} {\overrightarrow{\e{i}}}} {\t{n}}}
             {\filterset {\topprop{}} {\topprop{}}}
             {\emptyobject}
             {\topprop{}}}

@@

\begin{figure*}
$$
\begin{altgrammar}
{\javaspecial {\JavaField {\class{}}}} &=& {\javatotc {\class{}} {\true{}}}
\\
{\javaspecial {\JavaMethod {\overrightarrow{{\class{1}}}} {\class{2}}}} &=& 
    {\Arrow {\overrightarrow {\hastype {\x{}} {\javatotc {\class{1}} {\false{}}}}}
            {\javatotc {\class{2}} {\true{}}}
            {\filterset {\topprop{}} {\topprop{}}}
            {\emptyobject{}}
            {\topprop{}}}
\\
{\javaspecial {\JavaCtor {\overrightarrow{{\class{1}}}} {\class{2}}}} &=& 
    {\Arrow {\overrightarrow {\hastype {\x{}} {\javatotc {\class{1}} {\false{}}}}}
            {\javatotc {\class{2}} {\false{}}}
            {\filterset {\topprop{}} {\topprop{}}}
            {\emptyobject{}}
            {\topprop{}}}
\\\\

{\javatotc {\class{}} {\true{}}} &=& {\Union {\Nil} {\class{}}}
\\
{\javatotc {\class{}} {\false{}}} &=& \class{}
\\
{\javatotc {\JavaArray {\class{}}} {\true{}}} &=& {\Union {\Nil} {\javatotc {\ArrayTwo {\Bottom} {\class{}}} \true{}}}
\\
{\javatotc {\JavaArray {\class{}}} {\false{}}} &=& {\javatotc {\ArrayTwo {\Bottom} {\class{}}} \false{}}

@@

We currently do not support Java generics

>>>>

\title{Typed Clojure}
\subtitle{An Optional Type System for Clojure, based on Typed Racket}

\authorinfo{Ambrose Bonnaire-Sergeant}
           {University of Western Australia}
           {abonnairesergeant@gmail.com}
\authorinfo{Rowan Davies}
           {University of Western Australia}
           {rowan.davies@uwa.edu.au}

@@

\begin{abstract}
Clojure is a dynamically typed language hosted on the Java 
Virtual Machine.
Typed Racket is a valuable starting point for
a gradual type system that targets Clojure.
Building a similar type system for a new language gives the
designer some flexibility to repurpose and extend features.
This paper gives an overview of Typed Clojure, concentrating
on the extensions and differences from Typed Racket. We also
show where Typed Racket's features were particularly useful
for type checking non-trivial Clojure idioms.
\end{abstract}

@@

\section{Clojure with static typing}

% current situation

Programming is an error prone activity, especially in the presence of
multiple languages working together. Over the past decade, it has become increasingly
frequent to extend existing dynamically typed languages with optional type systems.
The guarantees that a sound type system provides can be the difference between
choosing a language that supports a type system or one that does not.

% presenting clojure

Clojure is a dynamically typed language built to run on the Java Virtual Machine (JVM)
which has recently gained popularity as an alternative JVM language.
It offers the flexibility of a Lisp dialect, including Common Lisp-style macros,
emphasising a functional style via a standard library of immutable datastructures. 
Clojure has good interoperability with existing Java code, allowing programmers to
use existing Java libaries without leaving Clojure.

%problem?

Clojure

@@

% list the delta from Typed Racket

@@

Typed Clojure is a gradual type system for Clojure. It is designed
to type check normal Clojure code by adding annotations. It is implemented
as a library, and can be seamlessly included in any Clojure project; no
separate compiler or language is needed.

@@

\begin{lstlisting}
(Fn [Integer -> Integer]
    [Number -> Number])
\end{lstlisting}

@@

\section{A Formal Model of Typed Clojure}

We describe our type system in a similar way to Typed Racket (\citet{TF10}).
The main judgement 
{\judgement {\Gamma} {\hastype {\e{}} {\t{}}} 
  {\filterset {\thenprop {\prop{}}}
              {\elseprop {\prop{}}}}
            {\object{}} {\flowprop {\prop{}}}}
says expression \e{} is of type \t{} in environment $\Gamma$, with 
then proposition {\thenprop {\prop{}}}, else proposition {\elseprop {\prop{}}},
object \object{} and flow proposition {\flowprop {\prop{}}}.

@@

Typed Clojure supports f-bounded polymorphism and higher-rank
types.
% why include them?
% examples, involving generic type operators for monads & conduits
% useful for generic programming, similar to Hashell type classes

@@

\begin{figure*}
$$
\begin{altgrammar}
  \expd{}, \e{} &::=& \x\ \alt {\comb {\e{}} {\e{}}} %\alt {\abs {\x a} {\e{}}}
                      \alt \true{} \alt \false{} \alt \nil{}
                      \alt \class{}
                      \alt {\tryexp {\e{}} {\class} {\x} {\e{}} {\e{}}}
                      \alt {\ifexp {\e{}} {\e{}} {\e{}}}
                      \alt
                      {\doexp {\e{}} {\e{}}}
                      \\ &\alt& 
                      {\fnexp {\overrightarrow{\hastype {\x{}} {\t{}}}} {\e{}}}
                      \alt {\letexp {\x{}} {\e{}} {\e{}}}
                &\mbox{Expressions} \\ \\
  \s{}, \t{}    &::=& \Top \alt (\class{}\ \overrightarrow{\t{}})
                      \alt {\Value \v{}}
                      \alt {\Unionsplice {\overrightarrow{\t{}}}}
                      \alt (\cap\ \overrightarrow{\t{}})
                      \alt (\Not\ \t{})
                      \alt {\All {\overrightarrow{\x{}}} {\t{}}}
                      \alt {\TFn {\overrightarrow{\x{}}} \t{}}
                      \alt {\Arrow {\overrightarrow{ \hastype {\x{}} {\t{}}}}
                                   {\t{}}
                                   {\filterset {\prop{}} {\prop{}}}
                                   {\object{}}}
                      \\ &\alt&
                      {\ArrayTwo {\t{}} {\t{}}}
                      \alt {\HMapc {TODO}}
                      \alt {\HMapp {TODO} {TODO}}

                &\mbox{Types} \\ \\
  \prop{}       &::=& {\isprop {\t{}} {\path {\pathelem{}} {\x{}}}}
                      \alt {\notprop {\t{}} {\path {\pathelem{}} {\x{}}}}
                      \alt {\impprop {\prop{}} {\prop{}}}
                      \alt {\andprop {\prop{}} {\prop{}}}
                      \alt {\orprop {\prop{}} {\prop{}}}
                      \alt \topprop{}
                      \alt \botprop{}
                &\mbox{Propositions} \\
  \object{}     &::=& {\path {\pathelem{}} {\x{}}}
                      \alt \emptyobject{}
                &\mbox{Objects} \\
  \pathelem{}   &::=& \overrightarrow{\pesyntax{}}
                &\mbox{Paths} \\
  \pesyntax{}   &::=& \classpe{} \alt \keype{k}
                &\mbox{Path Elements} \\ \\
                % TODO objects with path
%  \c{} &::=& \assoc{} \alt \dissoc{} \alt \get{}
%              &\mbox{Constants}
  \Gamma       &::=& \overrightarrow{\prop{}}
               &\mbox{Environments}
\end{altgrammar}
$$
\caption{Syntax of Terms, Types, Propositions, and Objects}
\end{figure*}

\begin{figure*}
\begin{mathpar}
\infer [T-Var]
{ \inpropenv {\propenv{}} {\isprop {\x{}} {\t{}}}}
{ \judgementfour {\propenv{}} {\hastype {\x{}} {\t{}}} {\filterset {{\notprop {\falsy} {\x{}}}} {{\isprop {\falsy} {\x{}}}}}
             {\x{}}}

\infer [T-True]
{}
{ \judgementfour {\propenv{}} {{\hastype {\true{}} {\True{}}}} {{\filterset {\topprop{}} {\botprop{}}}} {\emptyobject{}} }

\infer [T-False]
{}
{ \judgementfour {\propenv{}} {{\hastype {\false{}} {\False{}}}} {\filterset {\botprop{}} {\topprop{}}} {\emptyobject{}} }

\infer [T-Nil]
{}
{ \judgementfour {\propenv{}} {{\hastype {\nil{}} {\Nil{}}}} {\filterset {\botprop{}} {\topprop{}}} {\emptyobject{}} }

\infer [T-Do]
{ \judgementfour {\propenv{}} {\hastype {\e1} {\t1}} {\filterset {\thenprop {\prop{1}}} {\elseprop {\prop1}}} {\object{1}} 
\\\\
  \judgementfour {\propenv{}, {\orprop {\thenprop {\prop{1}}} {\elseprop {\prop{1}}}}}
                 {\hastype {\e{}} {\t{}}} {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}} 
                 {\object{}}
  }
{ \judgementfour {\propenv{}} {\hastype {\doexp {\e1} {\e{}}} {\t{}}} {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}} {\object} 
             }

\infer [T-If]
{ \judgementfour {\propenv{}} {\hastype {\e1} {\t{1}}} {\filterset {\thenprop {\prop{1}}} {\elseprop {\prop{1}}}}
                 {\object{1}} 
             \\\\
  \judgementfour {\propenv{}, {\thenprop {\prop{1}}}}
                 {\hastype {\e2} {\t{}}} {\filterset {\thenprop {\prop{2}}} {\elseprop {\prop{2}}}}
                 {\object{}}
  \\\\
  \judgementfour {\propenv{}, {\elseprop {\prop{1}}}}
                 {\hastype {\e3} {\t{}}} {\filterset {\thenprop {\prop{3}}} {\elseprop {\prop{3}}}}
                 {\object{}}
             }
{ \judgementfour {\propenv{}} {\hastype {\ifexp {\e1} {\e2} {\e3}} {\t{}}} 
                 {\filterset {\orprop {\thenprop {\prop{2}}} {\thenprop {\prop{3}}}} 
                             {\orprop {\elseprop {\prop{2}}} {\elseprop {\prop{3}}}}}
                 {\object{}} }

\infer [T-Let]
{ \judgementfour {\propenv{}} {\hastype {\e{1}} {\s{}}} {\filterset {\thenprop {\prop{1}}} {\elseprop {\prop{1}}}}
             {\object{1}}
  \\\\
  \judgementfour {\propenv{}, {\isprop {\s{}} {\x{}}}, 
              {\impprop {\notprop {\falsy} {\x{}}} {\thenprop {\prop{1}}}}, 
              {\impprop {\isprop {\falsy} {\x{}}} {\elseprop {\prop{1}}}}}
             {\hastype {\e{}} {\t{}}} {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
             {\object{}} 
             }
{ \judgementfour {\propenv{}} {\hastype {\letexp {\x{}} {\e{1}} {\e{}}} {\t{}}}
             {\replacefor {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
                          {\object{1}}
                          {\x{}}}
             {\replacefor {\object{}} 
                          {\object{1}}
                          {\x{}}}
             }

\infer [T-Throw]
{ \judgementfour {\propenv{}} {\hastype {\e{}} {\Throwable}} {\filterset {\thenprop {\prop{}}} {\elseprop {\prop{}}}}
                 {\object{}}
}
{ \judgementfour {\propenv{}} {\hastype {\throwexp {\e{}}} {\Bottom}} {\filterset {\botprop} {\botprop}}
             {\emptyobject{}} }

\infer [T-App]
{ \judgementfour {\propenv{}} {\hastype {\e{}} {\Arrow {\hastype {\x{}} {\s{}}}
                                                       {\t{}}
                                                       {\filterset {\thenprop {\prop{f}}}
                                                                   {\elseprop {\prop{f}}}}
                                                       {\object{f}}}}
                {\filterset {\thenprop {\prop{}}}
                            {\elseprop {\prop{}}}}
                {\object{}}
  \\\\
  \judgementfour {\propenv{}}
                 {\hastype {\ep{}} {\s{}}}
                 {\filterset {\thenprop {\propp{}}}
                             {\elseprop {\propp{}}}}
                 {\objectp{}} 
}
{ \judgementfour {\propenv{}} {\hastype {\appexp {\e{}} {\ep{}}}
                                        {\replacefor {\t{}}
                                                     {\objectp{}}
                                                     {\x{}}}}
                 {\replacefor {\filterset {\thenprop {\prop{f}}}
                                          {\elseprop {\prop{f}}}}
                              {\objectp{}}
                              {\x{}}}
                 {\replacefor {\object{f}}
                              {\objectp{}}
                              {\x{}}}
}

\infer [T-Abs]
{}
{ TODO }

\infer [T-Try]
{}
{ TODO }

\infer [T-New]
{ TODO }
{ \judgementfour {\propenv{}} {\hastype {\newexp {\class{}} {\overrightarrow{\e{i}}}} {\t{n}}}
             {\filterset {\topprop{}} {\botprop{}}}
             {\emptyobject}
             }

\infer [T-Field]
{ TODO }
{ }

\infer [T-Method]
{ TODO }
{ }

\infer [T-KwInvoke]
{}
{ TODO }

\infer [T-HMapInvoke]
{}
{ TODO }

\infer [T-HMapFull]
{}
{ TODO }

\infer [T-HMapAbsent]
{}
{ TODO }

\infer [T-HMapBoth]
{}
{ TODO }

\infer [T-Defmulti]
{}
{ TODO }

\infer [T-Defmethod]
{}
{ TODO }

\infer [T-InstanceOf]
{}
{ TODO }

\infer [T-Count]
{}
{ TODO }

\infer [T-Nth]
{}
{ TODO }

\infer [T-Class]
{}
{ TODO }

\infer [T-Subsume]
{}
{ TODO }

@@

\begin{figure*}
$$
\begin{altgrammar}
{\javaspecial {\JavaField {\class{}}}} &=& {\javatotc {\class{}} {\true{}}}
\\
{\javaspecial {\JavaMethod {\overrightarrow{{\class{1}}}} {\class{2}}}} &=& 
    {\Arrow {\overrightarrow {\hastype {\x{}} {\javatotc {\class{1}} {\false{}}}}}
            {\javatotc {\class{2}} {\true{}}}
            {\filterset {\topprop{}} {\topprop{}}}
            {\emptyobject{}}
            }
\\
{\javaspecial {\JavaCtor {\overrightarrow{{\class{1}}}} {\class{2}}}} &=& 
    {\Arrow {\overrightarrow {\hastype {\x{}} {\javatotc {\class{1}} {\false{}}}}}
            {\javatotc {\class{2}} {\false{}}}
            {\filterset {\topprop{}} {\topprop{}}}
            {\emptyobject{}}
            }
\\\\

{\javatotc {\class{}} {\true{}}} &=& {\Union {\Nil} {\class{}}}
\\
{\javatotc {\class{}} {\false{}}} &=& \class{}
\\
{\javatotc {\JavaArray {\class{}}} {\true{}}} &=& {\Union {\Nil} {\ArrayTwo {\Bottom} {\javatotc {\class{}} \true{}}}}
\\
{\javatotc {\JavaArray {\class{}}} {\false{}}} &=& {\ArrayTwo {\Bottom} {\javatotc {\class{}} \true{}}}

@@

We currently do not support Java generics
% because generics don't "exist" on the JVM
% clojure also doesn't use them
